%\documentclass[12pt]{article} %%% required

\documentclass[letterpaper,12pt]{article}
\usepackage{epsfig,colortbl}
\usepackage{amssymb}
\textwidth=6.5in
\textheight=9.5in
\topmargin=-0.75in
\oddsidemargin=0.0in
\evensidemargin=0.0in
\thispagestyle{empty} % no page numbers

\usepackage{times}
\usepackage{psfig}

\input /Users/raw/papers/texinit.latex

\begin{document} %%% required

\n {\Large \textbf{Final Exam --- AST 422 --- Spring 2012}} \hfill {\small Last
updated: Apr 16, 2012}

\bn {\bf DUE: May 2, 10:00 am, original in my mailbox in room PSF-686, plus
xerox in TA's mailbox.}

\bn {\kvbgbf Rules}:\ In the allotted time, please complete 6 questions of your
choosing from the set of 24 listed below. To make sure that all topics are
covered fairly, I ask you that you select at least a minimum number of
questions from each of the three main groups, as indicated below. So please
read all questions before you start. [E.g., you may select to do all of 
question A1 in category A, all of B1, B2 and B3 in category B, and all of C7
and C8 in category C, or any other permitted combination for a total of 6
questions in categories A, B, C.]

\mn All questions have equal weight, to be roughly equally distributed among
each of the sub-questions. Hence, out of a total of 150 points, each of the 6
questions that you choose will be worth 25 points, and each sub-question 
typically 5--8 points (for typically 5-3 sub-questions per question). 

\mn As Always: Brevity is as much appreciated as the correct answer. Please do
not give me more than 2--3 pages for each main question, if at all possible.

\mn NOTE: $H_0$ indicates the current value of the Hubble constant throughout, 
$q_0$ the current value of the deceleration parameter, $\Omega_0$ or 
$\Omega_m$ the current value of the density parameter, and $t_0$ the current 
age of the Universe. For some questions you must assume a cosmology, I 
suggest we all use $H_0 = 73\,km\,s^{-1}\,Mpc^{-1}$, $\Omega_m = 0.24$, 
$\Omega_{\Lambda} = \Lambda = 0.76$, unless given otherwise. In most FRW type
questions, I will assume $\Omega_{\Lambda} = 0$ for simplicity, so the 
answers can be found analytically -- in other questions we will use the 
``real'' cosmology $H_0 = 73$, $\Omega_m = 0.24$, $\Omega_{\Lambda} = \Lambda
= 0.76$. If you don't like solving these questions for $\Omega_{\Lambda} = 0$
you may solve them for $\Omega_{\Lambda} = \Lambda = 0.76$, but then you have
to use the numerical integration path. You may solve the questions for either
$\Omega_{\Lambda}$, but please tell us upfront which $\Omega_{\Lambda}$ you
have chosen.\\

\ve

\n\cl{\kvbgbf Group A: Select 1--2 Questions}\\

\vspace{-0.7cm} 
\begin{itemize}

\item [A1)] Consider the following 5 galaxies: A with $v = 5000\,km/s$ at 
distance $R = 100\,Mpc$; B with $v = 10,000\,km/s$ and $R = 200\,Mpc$; C with
$v = 5000\,km/s$ and $R = 50\, Mpc$; D with $v = 10,000\,km/s$ and $R =
100\,Mpc$; and E with at $z = 1.5$ with a measured age of $3.5\sim4.0\,Gyr$ at
that redshift.

\item [A1.a)] If your Universe only consists of galaxies A and B, what would
you conclude about the value of the Hubble constant and its error?

\item [A1.b)] If your Universe only consists of galaxies C and D, what would
you conclude about the value of the Hubble constant and its error?

\item [A1.c)] If your Universe consisted of galaxies A, B, C, and D, what
would you derive as the most likely value of the Hubble constant, if you also
knew that the distance measurements to the galaxies were uncertain by at least
35\%?

\item [A1.d)] If your Universe consisted of galaxies A, B, C, and D, what would
you be able to conclude about large scale motions and about the Hubble
expansion ($H_0$), if you knew that HST had measured their distances to better
than 15\% accuracy? How large could large-scale streaming motions be in such a
Universe? 

\item [A1.e)] Compute the age of the Universe if $H_0 = 73\,km/s/Mpc$ and
$\Omega = 0$. Roughly what would this age be for $\Omega_m = 0.24$, $1$, and
$2$ ($\Omega_{\Lambda} = 0$ in all cases)? Please sketch the expansion or
scale-factor of the Universe as a function of cosmic time for these different
values of $\Omega_m$. For which value of $\Omega_m$ do you get the oldest
Universe?

\item [A1.f)] Galaxy E is at a redshift $z = 1.5$, where the Universe was only
25\% of its current age (for $\Omega_m = 1$, $\Omega_{\Lambda} = 0$), and was
measured to have stellar population age of 3.5--4.0 Gyr at that redshift.
Discuss what problems this poses for certain values of $H_0$. How does this
compare to the globular cluster age problem (assuming the same values of $H_0$
and $\Omega_m$)? If $\Omega_m=1$, what values of $H_0$ would be implied by
galaxy E? Is this consistent with the most recent $H_0$ measurements? What
other solutions to this problem can you think of?\\

\item [A2.a)] Give a brief discussion (telegram style, no more than two 
pages) about what the major uncertainties are in determining the Hubble 
constant. Discuss in particular how Large Scale Structure can and has affected
the value of $H_0$.

\item [A2.b)] Out to what distances/on what scales do you know for sure that
the Universe is not homogeneous and isotropic, and at what distances/on what
scales should the Universe be homogeneous and isotropic? Document your answer.

\item [A2.c)] Describe briefly the differences and merits of the Cold Dark 
Matter (CDM) models of galaxy formation. Discuss briefly the major 
observational pieces of evidence in favor of/against CDM and in favor 
of/against HDM.\\

\item [A3.a)] Derive from scratch a period - Luminosity relation for 
Cepheids, using what you know about stellar structure (polytrope models etc), 
the stellar mass-luminosity relation, the relation between bolometric 
absolute magnitude and $M_V$, and the Hertzsprung Russell diagram. Give the
relevant intermediate steps, and derive: 
\\ $\log P = (0.5 h +0.3)\cdot M_V + (0.5hd + 0.3d + 3a)\cdot (B-V) + const$

\item [A3.b)] Discuss this result. What does each term mean, and how does it
apply to the distance scale problem?

\item [A3.c)] Approximately how is the zero-point in the Cepheid distance 
scale affected by dust? How does dust quantitatively affect the zero-point in
the Cepheid distance scale? How do (HST) distance measurements that use 
Cepheids attempt to correct for this?

\item [A3.d)] Derive for a simply rotating spiral galaxy disk, using Kepler's
laws, the Tully-Fischer relation: Luminosity $L$ proportional to $(\Delta
v)^4$, where $v$ in km/sec. Discuss your assumptions, how does this apply to
the distance scale problem?\\

\item [A4)] Describe briefly FOUR DIRECT methods for the determination of
$H_0$. Describe for each method roughly how it works, what the \% 
uncertainties are, and what the major causes for these uncertainties are. 
When describing these four methods, discuss at least the following:

\item [A4.a)] A method based on standard candles.

\item [A4.b)] A method based on standard rods.

\item [A4.c)] A method based on a physical model of hot gases.

\item [A4.d)] A method based on the effects of large masses on the path of 
light rays.\\

\item [A5)] Describe how the determination of the Hubble constant may be 
affected by: 

\item [A5.a)] The Malmquist bias. You may assume a Schechter or a Gaussian 
galaxy LF.

\item [A5.b)] Significant Large Scale Structure in the direction in which you
are measuring $H_0$.

\item [A5.c)] Significant large scale streaming motions (or a ``Great 
Attractor'') in the galaxy distribution. \\

\item [A6)] Define a problem of your choosing in nearby extragalactic 
astronomy and cosmology (topics: distance scale, large scale structure, etc.),
and solve it. You will be graded on this question for how well you pose the 
problem (25\%), how relevant it is to cosmology (25\%), and how well you 
solve it (50\%).\\

\end{itemize}

\ve

\n\cl{\kvbgbf Group B: Select 2--3 Questions}\\

\vspace{-0.7cm} 
\begin{itemize}

\item [B1.a)] Show that for all world models with $\Omega_{\Lambda}=0$, we 
have essentially $q_0=1/2$ or $\Omega_m=1$ shortly after creation, i.e. when 
$R(t) \rightarrow 0$ for $t \rightarrow 0$. Show that this is the case NOT ONLY
for $\kappa=0$, but also for $\kappa=-1$ and $\kappa=+1$. Discuss the
relevance of this result. (You may assume reasonable results from previous
homework questions. HINT: this issue is related to the flatness problem in
inflation). 

\item [B1.b)] Sketch the look-back time $\tau (z)$ for $\Omega_m=0$, $0.1$,
$1.0$, $2.0$. 

\item [B1.c)] Show that the Age of the Universe:\\ 
\vspace{0.2cm} $t_0 = \frac{1.0}{H_0}$ for $\Omega_m = 0.0$\\ 
\vspace{0.2cm} $t_0 = \frac{2}{3H_0}$ for $\Omega_m = 1.0$\\ 
\vspace{0.2cm} $t_0 = \frac{0.577}{H_0}$ for $\Omega_m = 2.0$\\ 
assuming $\Omega_{\Lambda}=0$ in all cases for simplicity.\\

\item [B2.a)] Assume that the cosmic time can be parametrized as function of
the ``epicyclic'' angle $\theta$, which is a monotonic function of cosmic 
time, and assume that:

$COS(\theta_0) = (2 - \Omega_m)/\Omega_m$\\
$COS(\theta_1) = (z + COS(\theta_0))/(1+z)$,

where $COS\equiv \cos$ for $\kappa=+1$ ($\Omega_m=2$ or $q_0=1$; 
$\Omega_{\Lambda}=0$) and $\equiv \cosh$ for $\kappa=-1$ ($\Omega_m=0$ or 
$q_0=0$; $\Omega_{\Lambda}=0$). Show that the look-back time as function of the
total age of the Universe can be written as: 
\[
\frac{\tau (z)}{t_0} = 1 - \left[ \frac{\theta_1 - \sin(\theta_1)}{\theta_0 -
\sin(\theta_0)} \right]
\]
for $\kappa=+1$, and
\[
\frac{\tau (z)}{t_0} = 1 - \left[\frac{\sinh(\theta_1) -
\theta_1}{\sinh(\theta_0) - \theta_0}\right]
\] for $\kappa=-1$,

and derive the product $t_0 H_0$ in both cases. 

\item [B2.b)] For general values of $\Omega_m$ or $q_0$, show that the 
look-back time as function of redshift can be written as: 
\[
\tau (z) = \frac{q_o}{H_0 (2q_0-1)^{3/2}} \cdot [\theta_0-\theta_1
+\sin(\theta_1)-\sin(\theta_0)]
\] 
for $\kappa=+1$, and\\
\[
\tau (z) = \frac{q_0}{H_0 (1-2q_0)^{3/2}} \cdot [\theta_1-\theta_0 +
\sinh(\theta_0)-\sinh(\theta_1)]
\]
for $\kappa=-1$

\item [B2.c)] Derive the special cases $\tau(z) = \frac{2}{3Ho}
\frac{(1+z)^{3/2} - 1}{(1+z)^{3/2}}$ for $\Omega_m=1$ or $q_0=1/2$ (assuming
$\Omega_{\Lambda}=0)$ and $\tau (z) = \frac{z}{H_0 (1+z)}$ for $\Omega_m=0$ or
$q_0=0$ (assuming $\Omega_{\Lambda}=0$). Sketch $\tau (z)$ for relevant values
of $\Omega_m$ in the range $\Omega_m=0\sim2$ (or $q_0=0\sim1$).\\

\item [B3.a)] Using the Mattig equation (Longair eq 7.38), show that the
observed angular size $\theta$ (in arcsec) of an object of physical size $Y$,
or a proper length $Y$, (in kpc) at redshift $z$ goes as:\\ 
\[ \theta = k \frac{Y (1+z)}{R_0 r} \] 
where ($R_0 r$) is the distance of the object as seen by the observer, and $k$
is some arbitrary constant. 

\item [B3.b)] Express $\theta$ in units of ($\frac{Y H_0}{c}$) as a function
of redshift $z$ for the case $\Omega_m=0$ or $q_0=0$ and the case $\Omega_m=1$
or $q_0=0.5$ (assuming $\Omega_{\Lambda}=0$ in both cases). Use Longair eq
5.56, 5.57 or Ryden eq 7.36.

\item [B3.c)] Sketch $\theta (z)$ for the cases $\Omega_m=0$, $1$ and $2$ (or
$q_0=0$, $0.5$, and $1$), and determine for $\Omega_m=0$ and $1$ (or $q_0=0$
and $q_0=0.5$) the value of the redshift at which $\theta$ will appear the
smallest (assuming $\Omega_{\Lambda}=0$ in all cases). 

\item [B3.d)] What is the physical size scale of a CBR fluctuation of $10'$ 
diameter at $z=1100$ for $\Omega_m=0$, $1$ and $2$ (or $q_0=0$, $0.5$, and $1$;
$\Omega_{\Lambda}=0$ in all cases)? What is it for the cosmology 
$\Omega_m=0.3$ and $\Omega_{\Lambda}=0.7$? At what angular scales do you 
expect the next generation of microwave satellites PLANCK to see most 
structure in the CBR, in terms of $\Delta T/T$? Document your answer. \\

\item [B4)] In all of the following sub-questions, make drawings of the run 
of the relevant parameter with redshift for various values of $\Omega_m=0\sim2$
(or $q_0=0\sim1$; assuming $\Omega_{\Lambda}=0$) and discuss the results. If 
you wish, you may do this for the more complicated case of 
$\Omega_m+\Omega_{\Lambda}=1$ for various values of $\Omega_m$. Where
necessary, you may assume that objects have a power-law spectrum with spectral
index alpha, $\alpha$. 

\item [B4.a)] Discuss how the luminosity distance depends on redshift. 

\item [B4.b)] Discuss how the volume element available per unit redshift 
depends on redshift. Discuss at least two reasons why there may be a decline 
in the observed quasar and young galaxy density for $z > 2\sim3$. 

\item [B4.c)] Discuss how the observed integrated bolometric and 
monochromatic flux of an object of fixed standard spectrum depends on 
redshift. 

\item [B4.d)] Discuss how the observed bolometric and monochromatic surface 
brightness of an object of fixed standard spectrum depends on redshift. 

\item [B4.e)] Discuss how the angular size of an object of fixed metric 
linear size depends on redshift. \\

\item [B5.a)] Show why the matter density in the post-recombination Universe
scales as $(1+z)^3$, and why the radiation density has scaled since then --
and even before then -- as $(1+z)^4$. What does this mean for the shape of the
Cosmic Background Radiation spectrum and for the CBR temperature as function
of $(1+z)$? 

\item [B5.b)] Show qualitatively (i.e. never mind the exact physical 
constants) why the Jeans mass of a spherical clouds of hot matter goes as 
$T^{-3}$ before recombination, and as $T^{3/2}$ after recombination. Discuss 
what -- as a consequence -- the typical masses are of gas clouds that first 
fragment into stars or groupings of stars in the early Universe. 

\item [B5.c)] Discuss briefly why the faint galaxy two-point correlation 
function has an amplitude that is so low. What does this tell you about the 
redshift distribution of the faint blue galaxy population and about the 
clustering of faint galaxies (and its evolution, or lack thereof). \\

\item [B6.a)] A light source in a flat universe has a redshift $z$ when 
observed at a time $t_0$. Show that the observed redshift changes with time 
at a rate of:\\
\[ \frac{dz}{dt_0} = H_0 (1+z) - H_o (1+z)^{3(1+\omega)/2} \]
where $\omega = 0$ for matter+vacuum-energy dominated universes (you may 
assume $\omega=0$). For what values of $\omega$ does the redshift decrease 
with time? For what values of $\omega$ does the redshift increase with time?

\item [B6.b)] Suppose you are in a flat, matter-only universe that has a 
Hubble constant\\ $H_0=73\,km/sec/Mpc$. You observe a galaxy wth redshift 
$z=1.000000$. How long will you have to be observing this galaxy from Earth 
to see its redshifts change by one part in $10^6$? [Use th results from 6.a]\\

\item [B7)] Define a problem of your choosing in cosmology and the early 
universe (topics: expansion, FRW models, geometry, observational distinction 
between the various cosmological models, etc), and solve it. You will be 
graded on this question for how well you pose the problem (25\%), how 
relevant it is to cosmology (25\%), and how well you solve it (50\%).\\

\end{itemize}

\ve

\n\cl{\kvbgbf Group C: Select 2--3 Questions}\\

\vspace{-0.7cm}
\begin{itemize}

\item [C1.a)] For a population of cosmic radio sources emitting radiation 
with an assumed constant luminosity or radio power $P$ (in $W/Hz$) and 
constant space density (in $N_0/Mpc^3$), derive for the Euclidean case the 
source count $N(S)$, where $N$ is the number of sources (per ster rad) with 
observed flux density greater than some value $S$ (measured in $Jy=10^{-26}
W/Hz/m^2$). 

\item [C1.b)] What is the logarithmic slope of the source counts in flux 
units ($S$), and in magnitude units [$m = -2.5 \log (S) + const$] for this 
Euclidean case? Sketch this together the radio source counts observed by Ryle
et al. and the optical galaxy counts by many recent observers. What are the
actual observed slopes at the bright end of the counts in both radio and 
optical? How do you interpret the slopes being different from question a? 
Discuss all possible reasonable answers. 

\item [C1.c)] At approximately what magnitude do the B-band galaxy counts 
make a significant change in slope (slope changing from approximately what 
value to what other value)? Calculate schematically the contribution of each 
magnitude bin of the B-band galaxy counts to the Extragalactic Background 
Light (EBL) for $10<B<30 mag$. At which magnitude do the B-band galaxy counts
contribute maximally to the EBL? How can this be interpreted in terms of their
redshift distribution, and what does this imply for galaxy 
formation/evolution scenarios?\\

\item [C2.a)] Discuss why it was -- in hindsight -- not surprising that in 
the deepest available HST fields (~50 hr exposures in both HDF's), faint 
galaxies were seen down to $V$, $I=30 mag$ which are generally quite compact, 
with scale-lengths $r_{hl} = 0.1\sim0.3''$. Discuss why the most compact 
objects are relatively easy to detect with HST, what observational selection 
effects have to be overcome, and how their flux depends on the various 
factors of $(1+z)$. 

\item [C2.b)] The Next Generation Space Telescope (NGST) is an instrument 
currently planned for launch in 2007. It will have an 8 meter (segmented 
mirror) and a rich array of optical and infrared camera's and spectrographs. 
Estimate how faint this telescope can see in a 1000 hr exposure (Hint: 
compare to what the 2.4 meter HST can do with its 50 hr exposures in the HDF
in V and I). 

\item [C2.c)] Assume the IR detectors are good enough that similar 
sensitivities can be obtained in the near-IR JHK bands. For $H_0=73$ and 
$\Omega_m=0.24$, compute out to which redshift this telescope can see AGN 
($M_V=-25$ and $r_{hl}<0.1\,pc$), $L^{\ast}$ galaxies ($M_V=-22$ and 
$r_{hl}=5\,kpc$), sub-galactic objects ($M_V=-17$ and $r_{hl}=0.1\,kpc$),
supernovae ($M_V=-19$), and globular clusters ($M_V=-10$ and $r_{hl}=1\,pc$). 
Assume for simplicity that all these objects are young and therefore blue 
with flat power-law spectra ($\alpha=0$ or $1$), so that K-corrections are 
straightforward. [Note: You need to get the answer to question b right to 
answer c correctly.]

\item [C2.d)] What are likely the only objects that the NGST will be able to
see at $z>10$? Argue your answer and mention at least two reasons why this
category of object is favored above all the others. (You may assume that AGN
form after at least one galactic free fall time has passed).\\

\item [C3.a)] Consider a stellar population of $10^{10}\,M_{\odot}$ in total, 
all born at the same time in the very early history of a galaxy, and following
a Salpeter IMF between $0.08\,M_{\odot}<M<100\,M_{\odot}$ -- with no stars 
forming outside this range. Roughly how large a fraction of the stars in this 
galaxy will go off in Supernovae? How many stars is this? How often do they go
off as SNe? 

\item [C3.b)] Roughly when did this Supernovae phase happen during the 
evolution of the galaxy? What redshift range does this roughly correspond to,
if the entire galaxy formed within one free fall time after the Big Bang? 

\item [C3.c)] Sketch qualitatively how the various stellar populations that
develop in the HR diagram (i.e. the MS, RGB, HB, AGB, P-AGB, etc) during the 
evolution of this galaxy contribute to its total galaxy light in the Johnson 
filters UBVRI+JHK, as well as in the far UV (around $1500$ and $2500 \AA$). 
Briefly discuss which component dominates each filter and why. 

\item [C3.d)] Compute the probability that this galaxy is seen with a 
supernovae, if it were discovered at very high redshift by the HST or NGST. 
[You need to know the answer to question a and b to answer d].\\

\item [C4.a)] Discuss briefly how the integrated optical/near-IR spectra of 
galaxies depend on the age of their dominant stellar population. Make a 
schematic drawing of their spectra. How does this affect the K-correction we 
derive for such galaxies? 

\item [C4.b)] Discuss briefly how the integrated optical/near-IR spectra of
galaxies depend on the past star-formation history. Make a schematic drawing of
their spectra for various assumed (cartoon-like) star-formation histories. 

\item [C4.c)] Discuss how the answers of questions a and b are connected: 
what is the physical cause for this connection? How can we best measure 
galaxy ages?\\

\item [C5.a)] Sketch and discuss how the galaxy counts determined as function
of Hubble type enable us to constrain the galaxy luminosity function (LF) and
its evolution with cosmic epoch.

\item [C5.b)] What is the major uncertainty in these exercises? How can we 
break this degeneracy by measuring the redshifts or the redshift distribution 
of faint galaxies? 

\item [C5.c)] Discuss evidence in favor of, and evidence against the premise
that faint blue galaxies are primarily lower-luminosity (i.e. $L<L^{\ast}$ in 
the LF) objects of irregular morphology and with sizes typically smaller than 
those of NGC galaxies seen today. What does this mean for the formation 
scenario of galaxies? \\

\item [C6)] You observe a quasar at a redshift of $z=5.0$ and determine that 
the observed light flux from the quasar varies on a timescale $\Delta t_0 = 3$
days. If the observed variation in flux is due to a variation in the intrinsic
luminosity of the quasar, what was the variation timescale $\delta t_e$ at the
time the light was emitted? For the light of the quasar to vary on a timescale
$\delta t_e$, the bulk of the quasar light must come from a region of physical
size $R \le R_{max} = c\,\delta t_e$ . What is the angular size of $R_{max}$
in the standard model? How would you go about observing this?\\

\item [C7)] In a flat Universe with $H_0=73 km/s/Mpc$, you observe a galaxy 
at a $z=6$. 

\item [C7.a)] What is the current proper distance to the galaxy, $d_p(t_0)$,
if the universe only contains radiation?

\item [C7.b)] What is $d_p(t_0)$ if the Universe only contains matter?

\item [C7.c)] What is $d_p(t_0)$ if the Universe only contains a cosmological
constant?

\item [C7.d)] What was the proper distance at the time the light was emitted,
$d_p(t_e)$, if the universe only contains radiation?

\item [C7.e)] What was the proper distance at the time the light was emitted, 
$d_p(t_e)$, if the universe only contains matter?

\item [C7.f)] What was the proper distance at the time the light was emitted,
$d_p(t_e)$, if the universe only contains a cosmological constant?\\

\item [C8)] The surface brightness SB of an astronomical object is defined as
its observed flux divided by its angular area: $SB = f/(\delta theta)^2$. 

\item [C8.a)] For a class of objects that are both standard candles and 
standard yardstick, what is SB as a function of redshift in the case $z<<1$?

\item [C8.b)] For a class of objects that are both standard candles and 
standard yardstick, what is SB as a function of redshift in the case $z>>1$?

\item [C8.c)] Would the SB of this class of object be useful way of 
determining the deceleration parameter qo? Why or why not?

\item [C8.d)] Discuss the grave cosmological consequences of your answer in
C-8.b. Why do you think observations with HST at $z>>5$ are so hard? What do
you think objects that JWST will see at $z=15$ will look like to this 
telescope --- resolved or unresolved? Document your answer carefully.\\

\item [C9)] Define a problem of your choosing in galaxy formation and 
evolution (topics: galaxy formation, galaxy evolution, AGN evolution, 
structure formation, clustering evolution, the Intergalactic Medium etc), and
solve it. You will be graded on this question for how well you pose the 
problem (25\%), how relevant it is to cosmology (25\%), and how well you 
solve it (50\%). 

\end{itemize}

\end{document} %%% required
